---
title: "MSU - Week 3: Statistical models of Differential Expression"
date: "08/15/2015"
output: html_document
layout: topic
minutes: 20
---

# The mean-variance relationship (and what it means)

Let's start by loading the data:
```{r load}
load('bottomly_eset.RData')
```

Take a look at your environment, and you should find that three ExpressionSet objects have been loaded along with all the required libraries. _If you have errors with this, you may have to re-install the package. Please take a look at the [README](./README.md) for more details on requirements for this module._

The three datasets correspond to the full [Bottomly](http://www.ncbi.nlm.nih.gov/pubmed?term=21455293) dataset that is available on [ReCount](http://bowtie-bio.sourceforge.net/recount/), and two smaller datasets generated by randomly selecting samples from the full set. The full set contains data from two different mouse strains C57Bl and DBA, with 10 and 11 replicates, respectively. We created another dataset with 5 replicates for each strain (n = 10), and one with 2 replicates for each strain (n = 4). **How does the relationship between mean and variance change as you increase the number of replicates?**

```{r mean-variance, fig.align='center'}

# Plot 2 replicate dataset
eset <- bottomly.2reps
cpm.mat <- log(cpm(exprs(eset)))
mean.vec <- apply(cpm.mat, 1, mean)
sdvec <- apply(cpm.mat, 1, sd)
plot(mean.vec, sdvec, pch=".", main="2 replicates", ylab="sd", xlab="Average logCPM")

# Plot 5 replicate dataset
eset <- bottomly.5reps
cpm.mat <- log(cpm(exprs(eset)))
mean.vec <- apply(cpm.mat, 1, mean)
sdvec <- apply(cpm.mat, 1, sd)
plot(mean.vec, sdvec, pch=".", main="5 replicates", ylab="sd", xlab="Average logCPM")

# Plot 10 replicate dataset
eset <- bottomly.eset
cpm.mat <- log(cpm(exprs(eset)))
mean.vec <- apply(cpm.mat, 1, mean)
sdvec <- apply(cpm.mat, 1, sd)
plot(mean.vec, sdvec, pch=".", main="10 replicates", ylab="sd", xlab="Average logCPM")

dev.off()

```

Appropriate modeling of the mean-variance relationship in DGE data is important for making inferences about differential expressionIn RNA-Seq data, genes with larger average expression have on average larger observed variances across samples, that is, they vary in expression from sample to sample more than other genes with lower average expression. This phenomena of 'having different scatter' is known as data **heteroscedasticity**. In our figures, we observe this heteroscedasticity except that the logarithmic transformation counteracts this and overdoes the adjustment somewhat so that large log-counts now have smaller scatter than small log-counts. More, importantly we see the scatter tends to reduce as we increase the number of replicates.

# Statistical models for DE analysis
The low levels of replication rule out, for all practical purposes, distribution-free rank or permutation-based methods. The Poisson model doesn't work either -- it is a single parameter model with mean == variance and real data has more variance than Poisson can explain (which we will look at in our dataset). The Negative Binomial model is a good approximation, where the variability between replicates is modeled by the **dispersion parameter**. 

[edgeR](http://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf) and [DESeq2](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.pdf) are two very popular tools whcih implement differential analysis on the basis of the NB model. They offer overlapping functionality but differ crucially in the way dispersions are estimated. Accurate estimation of the dispersion parameter is critical for the statistical inference of differential expression.

## DESeq2
Both methods utilize the concept of “borrowing information from genes”; using the whole dataset to compute a common value, trend, or prior distribution for the dispersions, and then shrink individual gene-wise dispersion estimates toward this chosen anchor. With DESeq2 the first round of estimates are computed using maximum likelihood, and a curve is fit to provide an estimate of expected dispersion. Then the gene-wise dispersion estimates are shrunken toward the values predicted by the curve to obtain final
dispersion values. Look at the dispersion-mean plots below, What do you see?

```{r deseq2}
# Create DESeq2 datasets
dds <- DESeqDataSetFromMatrix(countData = exprs(bottomly.eset), colData = pData(bottomly.eset), design = ~ strain )
dds <- DESeq(dds)

dds.5rep <- DESeqDataSetFromMatrix(countData = exprs(bottomly.5reps), colData = pData(bottomly.5reps), design = ~ strain )
dds.5rep <- DESeq(dds.5rep)

dds.2rep <- DESeqDataSetFromMatrix(countData = exprs(bottomly.2reps), colData = pData(bottomly.2reps), design = ~ strain )
dds.2rep <- DESeq(dds.2rep)


# Plot dispersion estimates
plotDispEsts(dds.)
plotDispEsts(dds.5rep)
plotDispEsts(dds.2rep)
```

The final estimates for datasets with so few replicates result in values that are so far off from the initial estimates. This is because the strength of shrinkage depends (i) on an estimate of how close true dispersion values tend to be to the fit and (ii) on the degrees of freedom. _As the sample size decreases, the shrinkage increases in strength._ In the case of a 2 replicate dataset it seems that for many genes the dispersion is being well overestimated which may lower the rate of true detection.

